---
title: "The Relationship of Mathematics Scores to Reading Scores"
author: "Halle Prine"
date: "`r Sys.Date()`"
format: html
---

The *HSB2* data set consists of data from 200 randomly selected high school seniors. These data include five test scores, as well as demographic data. This report will show illustrate the relationship of mathematics scores to reading scores separately for each of three socioeconomic categories: low, medium, and high.

```{r message=FALSE, warning=FALSE, include=FALSE}
# Clear the environment
rm(list = ls())

# Load the libraries
library(here)

# Read in the data
hsb2 <- read.csv(here("Data", "hsb2.csv"))

# Create a separate matrix for each SES level that contains the five test scores
hsb2_low <- as.matrix(hsb2[hsb2$ses == 1, 7:11])
hsb2_med <- as.matrix(hsb2[hsb2$ses == 2, 7:11])
hsb2_high <- as.matrix(hsb2[hsb2$ses == 3, 7:11])

# Eliminate matrix row names
row.names(hsb2_low) <- NULL
row.names(hsb2_med) <- NULL
row.names(hsb2_high) <- NULL

# Create a list with these three matrices
hsb2_list <- list(Low = hsb2_low,
                  Medium = hsb2_med,
                  High = hsb2_high)

# Remove the temporary objects
rm(hsb2, hsb2_low, hsb2_med, hsb2_high)
```

```{r include=FALSE}
# Calculate correlations and p-values for each SES level
low_corr <- cor.test(hsb2_list[["Low"]][, "read"], hsb2_list[["Low"]][, "math"])
med_corr <- cor.test(hsb2_list[["Medium"]][, "read"], hsb2_list[["Medium"]][, "math"])
high_corr <- cor.test(hsb2_list[["High"]][, "read"], hsb2_list[["High"]][, "math"])
```

Below we have provided scatter plots Figure 1 to demonstrate the relationship between highschool student's reading and math score among 
socioeconomic statuses (i.e., Low, Medium, and High). Table 1 provides the exact correlation table we can see that SES shares a moderately strong positive relationship between reading and math scores among low SES (`r round(low_corr$estimate, 2)` , p = `r round(low_corr$p.value, 3)`), (r = `r round(med_corr$estimate, 2)`, p = `r round(med_corr$p.value, 3)`), and high SES (r = `r round(high_corr$estimate, 2)`, p = `r round(high_corr$p.value, 3)`).

*Table 1. Correlation Table Between Reading and Math Scores among Low, Medium, and High SES Status*

```{r echo=FALSE}

# Calculate correlations and p-values for each SES level
low_corr <- cor.test(hsb2_list[["Low"]][, "read"], hsb2_list[["Low"]][, "math"])
med_corr <- cor.test(hsb2_list[["Medium"]][, "read"], hsb2_list[["Medium"]][, "math"])
high_corr <- cor.test(hsb2_list[["High"]][, "read"], hsb2_list[["High"]][, "math"])

# Create a simple correlation table with significance
correlation_table <- data.frame(
  SES = c("Low", "Medium", "High"),
  Correlation = round(c(low_corr$estimate, med_corr$estimate, high_corr$estimate), 2),
  P_value = round(c(low_corr$p.value, med_corr$p.value, high_corr$p.value), 3),
  Significance = c(ifelse(low_corr$p.value < 0.05, "Significant", "Not Significant"),
                   ifelse(med_corr$p.value < 0.05, "Significant", "Not Significant"),
                   ifelse(high_corr$p.value < 0.05, "Significant", "Not Significant"))
)

# Print the table
print(correlation_table)

```

*Scatter Plots of Reading and Math Scores among Low, Medium, and High SES Status*

```{r echo=FALSE}
par(mfrow = c(1, 3))  # Set up 3 plots in one row

for (ses in names(hsb2_list)) {  
  scores <- hsb2_list[[ses]]  
  
  # Calculate the correlation between reading and math scores
  correlation <- cor(scores[, "read"], scores[, "math"])
  
  # Round the correlation value for cleaner display
  rounded_correlation <- round(correlation, 2)
  
  # Create the scatter plot for reading vs math
  plot(scores[, "read"], scores[, "math"],  
       xlab = "Reading Score", ylab = "Math Score",  
       main = paste("SES:", ses), pch = 16, col = "blue")  
  
  # Add the correlation text to the plot
  text(x = max(scores[, "read"]) - 10, y = max(scores[, "math"]) - 10, 
       labels = paste("Correlation:", rounded_correlation), col = "red", cex = 1.2)
  
  # Optionally print the correlation for each SES group
  print(paste("SES:", ses, "Correlation:", rounded_correlation))
}

par(mfrow = c(1, 1))  # Reset plot layout

```
