---
Author: "Halle Prine"
title: "EDRM 718 Exercise Set J"
format: html
---

*Put your name as the author of this notebook in the YAML, then complete each of the items below. Make sure that your notebook is neat and organized and that all your code chunks run without error, even after clearing the environment. (Six points per item)*

***

*Use **ggplot2()** for all items in this exercise set. Use the **hsb2** data for item 1 through item 6.*

```{r echo=FALSE}

library(ggplot2)
library(here)
here()

# Had to reset my Here library 
setwd("/Users/halleprine/Desktop/GitREPO/EDRM 718 HP - Spring")
getwd()
here::i_am("Data/hsb2.csv")


# I read the data and put it in an object called "school".
school <- read.csv(here("Data", "hsb2.csv"), header = FALSE, sep = ",") # Read the file
View(school)

# Columns not named correctly I had to go in and rename them and remove first row
names(school)[names(school)=="V1"]<-"ID" 
names(school)[names(school)=="V2"]<-"sex"
names(school)[names(school)=="V3"]<-"race"
names(school)[names(school)=="V4"]<-"ses"
names(school)[names(school)=="V5"]<-"schtyp"
names(school)[names(school)=="V6"]<-"prog"
names(school)[names(school)=="V7"]<-"read"
names(school)[names(school)=="V8"]<-"write"
names(school)[names(school)=="V9"]<-"math"
names(school)[names(school)=="V10"]<-"science"
names(school)[names(school)=="V11"]<-"socst"

school <- school[-c(1), ]

```


***

**1. Create a histogram of math scores.**

```{r echo=FALSE}
str(school$math)
school$math <- as.numeric(as.character(school$math))

as.tibble(schoo)

ggplot(school, aes(x = math)) +
  geom_histogram(binwidth = 5, fill = "blue", color = "black", alpha = 0.7) +
  labs(title = "Histogram of Math Scores",
       x = "Math score",
       y = "Frequency") +
  theme_minimal()

```

***

**2. Create a scatterplot to illustrate the relationship of math and science scores. Put math on the horizontal axis.**

```{r echo=FALSE}

school$science <- as.numeric(school$science)
school$math <- as.numeric(school$math)

ggplot(school, aes(x = math, y = science)) +
  geom_point(color = "blue") +
  geom_smooth(method = "lm", color = "red", se = FALSE) +  
  labs(title = "Scatterplot of Math vs. Science Scores",
       x = "Math Score",
       y = "Science Score") +
  theme_minimal()


```

***

**3. Create a scatterplot to illustrate the relationship of math and science scores. Put math on the horizontal axis. Use different colors to indicate the SES for each student in the plot. Put a linear regression line on your plot.**

```{r echo=FALSE}

ggplot(school, aes(x = math, y = science, color = factor(ses))) +
  geom_point() +
  geom_smooth(formula = y ~ x, method = "lm", se = FALSE)
  labs(title = "Scatterplot of Math vs. Science Scores with SES",
       x = "Math Score",
       y = "Science Scores",
       color = "SES") +
  theme_minimal()

```

***

**4. Create two faceted scatterplots to illustrate the relationship of math and science scores: one for private school students and one for public school students. Put math on the horizontal axis of each plot. Use different colors to indicate the SES for each student in the plots. Put a linear regression line on your plots.**

```{r echo=FALSE}

ggplot(school, aes(x = math, y = science, color = factor(ses))) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, aes(group = 1), color = "black") +
  facet_wrap(~ factor(schtyp, labels = c("Public", "Private"))) +
  labs(title = "Scatterplot of Math vs. Science Scores by School Type",
       x = "Math Score",
       y = "Science Scores",
       color = "SES") +
  theme_minimal()

```


***

**5. Create side-by-side boxplots to illustrate the differences in math scores of students from different SES groups. Include points for every student in your plot. Also plot the mean for each group.**

```{r echo=FALSE}

ggplot(school, aes(x = factor(ses, levels = c(1, 2, 3), labels = c("Low", "Medium", "High")), 
                    y = math, color = factor(ses, levels = c(1, 2, 3), labels = c("Low", "Medium", "High")))) +
  geom_boxplot(fill = "lightblue", color = "black", outlier.shape = NA) +  
  geom_jitter(width = 0.2, alpha = 0.6) +  
  stat_summary(fun = "mean", geom = "point", shape = 18, size = 3, color = "red") +  
  theme_minimal() +
  labs(title = "Differences in Math Scores by SES Group",
       x = "SES Group",
       y = "Math Score",
       color = "SES") +  # Ensures the legend title is correct
  scale_color_manual(values = c("red", "blue", "green")) +  
  theme(legend.position = "right")

```


***

**6. Create stacked bar plots to illustrate how SES groupings is related to the private-public school distinction. Plot proportions on the vertical axis.**

```{r echo=FALSE}

ggplot(school, aes(x = factor(schtyp, levels = c(1, 2), labels = c("Public", "Private")), 
                   fill = factor(ses, levels = c(1, 2, 3), labels = c("Low", "Medium", "High")))) +
  geom_bar(position = "fill") +  # 'fill' scales bars to proportions
  scale_y_continuous(labels = scales::percent_format()) +  # Convert y-axis to percentages
  labs(title = "Proportion of SES Groups in Public vs. Private Schools",
       x = "School Type",
       y = "Proportion",
       fill = "SES Group") +  # Legend title
  theme_minimal()

```
***

*Use the **mental health study* *data for item 7.*

***

**7. Use the mental health study data to graphically explore the relationship of life events and SES on mental health impairment. Your graph should illustrate whether SES moderates the influence of life events on mental health.**

```{r echo=FALSE}

library(readxl)

mental <- read_excel(here("Data", "Mental Health Study Data.xlsx"), sheet = 1)
View(mental)


mental%>%
  mutate(SES_cat = cut (SES, breaks = 2, labels = c("Low", "High"))) %>%
  ggplot(aes(x = LifeEvts, y = MentImp, color = SES_cat)) +
  geom_point() +
  geom_smooth(method = "lm", se = TRUE, linetype = "solid") + 
  labs(title = "Moderating Effect of SES on Life Events and Mental Health Impairement",
       x = "Life Events",
       y = "Mental Health Impairement") +
  scale_color_discrete(name = "SES")


```


***

*Use the **notes study**data for item 8.*

***

**8. Use the notes study data to graphically explore the relationship of method of note taking to later recall. Your graph should illustrate whether the type of lecture moderates the effect of note taking method on recall.**

```{r echo=FALSE}

notes_dta <- read_excel(here("Data", "Notes Data.xlsx"), sheet = 1)
View(notes)

ggplot(notes_dta, aes(x = notes, y = score, fill = lecture)) +
  geom_boxplot(alpha = 0.4, position = position_dodge(0.8), outlier.shape = NA) +  # Boxplots in the background
  stat_summary(fun = mean, geom = "point", aes(color = lecture), size = 3, position = position_dodge(0.3)) +  # Mean points
  stat_summary(fun = mean, geom = "line", aes(color = lecture, group = lecture), position = position_dodge(0.3)) +  # Connect means
  labs(title = "Effect of Note-Taking Method on Recall by Lecture Type",
       x = "Note-Taking Method",
       y = "Recall Score",
       fill = "Lecture Type",
       color = "Lecture Type") +
  theme_minimal()

```


***
