---
title: "EDRM 718 Exercise Set E"
Author: "Halle Prine
output:
  html_document:
    df_print: paged
---

*Put your name as the author of this notebook in the YAML, then complete each of the items below. Make sure that your notebook is neat and organized and that all your code chunks run without error, even after clearing the environment. (Six points per item)*

**For items 1 through 4 use the HSB2 data to illustrate how each of the four members of the *apply* family works. Do this by using the correct family member to accomplish each of the following tasks for the five test scores (not the entire data set). Make sure to use each family member for one, and only one, item.**
```{r include=FALSE}
# The "here" library will let me read the data from an individual folder.
library(psych)
library(here)

# I read the data and put it in an object called "legislators".

students <- read.csv(here("Data", "hsb2.CSV"), header = TRUE, sep = ",") # Read the file

# Examining the Variables for students 
View(students)
colnames(students)

```

***

**1. Put the standard deviation (SD) of each test as an element in a list of the five SDs.**

```{r include=FALSE}
# Select only the five test score columns 

test_scores <- students[, c("read", "write", "math", "science", "socst")]

# Compute SD using lapply() and store the results in a list 

sd_list <- lapply(test_scores, sd)

sd_list

```


***

**2. Create a vector of SDs for the five tests.**

```{r include=FALSE}
# Compute SD using sapply() and store the results in a vector

sd_vector <- sapply(test_scores, sd)

sd_vector

```

***

**3. Create a vector of the standard deviation of the five test scores for each student in the sample.**

```{r include=FALSE}
# Compute SD for each student (row-wise) using apply()

sd_student <- apply(test_scores, 1, sd)

sd_student

```


***

**4. Obtain SDs for the five tests by school type (i.e., public or private).**

```{r include=FALSE}
# Compute SDs for the five test scores grouped by school type
sd_school <- aggregate(test_scores, by = list(students$schtyp), FUN = sd)

# Rename the grouping column for clarity
colnames(sd_school)[1] <- "School_Type"

# Print the results
sd_school

```

***

**5. Using the list you created in item 1, store the reading standard deviation from this list in three new objects, *read_sd_1*, *read_sd_2*, and *read_sd_3* to demonstrate the use of these three operators: [ ], [[ ]], and $. Examine the class of each object and briefly explain why this is the class.**

```{r include=FALSE}
# Extract the SD for reading using different operators
read_sd_1 <- sd_list["read"]   # Using [ ]
read_sd_1 

read_sd_2 <- sd_list[["read"]] # Using [[ ]]
read_sd_2

read_sd_3 <- sd_list$read      # Using $
read_sd_3

# Examine the class of each extracted object
class(read_sd_1)  # Check class of read_sd_1
class(read_sd_2)  # Check class of read_sd_2
class(read_sd_3)  # Check class of read_sd_3
```

***

**In an early exercise set you created vectors and a data frame that included the names of some of the members of our class, email addresses, and the distance we each travel to our class meetings. For the exercises below you will use these same data for all students in our class. (At least one member from each of the previous groups will need to share the data for their group. You may use the Research Cafe, or other means if you prefer, to do this.) You will use this combined data set for the following items, though you should just include the current members of our class. The class member profiles is current, so you can use that if you need a reminder of who is currently in the class.**

***

**6. Create a data frame named *our_class* that includes these variables, in order: *first_name*, *last_name*, *email*, *distance*.**

```{r include=FALSE}
# Create vectors for each variable
first_name <- c("Adam", "Aishwanya", "Alison", "Beatrice", "Benny", "Brandi", 
                "Emily", "Gitanjali", "Halle", "Irene", "Jongwon", "Julia", 
                "Kaitlyn", "Kartik", "Matt", "Nestor", "Regin", "Saeed", 
                "Samin", "Sayema", "Suprja", "Unum")

last_name <- c("Brown", "Khemkar", "Scheide", "Quiroz", "Thomswarzecha",
               "Byers", "Lorbach", "Lall", "Prine", "Esu", "Lee", "Barnes",
               "Ramey", "Midha", "Marquiez auza","Urrea romero", "Dean", 
               "Ahmadi", "Ashjael", "Akter", "Arikumar", "Aslam")

email <- c("Adam.brown@moore.sc.edu", "akhemkar@email.sc.edu", 
           "scheidea@mailbox.sc.edu", "bnquiroz@email.sc.edu",
           "thomswab@email.sc.edu", "byersb@mailbox.sc.edu",
           "lohrbach@email.sc.edu", "glall@email.sc.edu",
           "hprine@email.sc.edu", "iesu@email.sc.edu",
           "lee1125@sc.edu", "jab30@mailbox.sc.edu",
           "ktramey@email.sc.edu", "kmidha@email.sc.edu",
           "mateom@email.sc.edu", "nfu@email.sc.edu",
           "regin@email.sc.edu", "ahmadis@email.sc.edu",
           "sashjaei@email.sc.edu", "sakter@email.sc.edu",
           "Srikumar@email.sc.edu", "uaslam@email.sc.edu")

distance <- c(12, 1, 5, 5, 2, 19, 1, 1, 7, 1, 19, 2, 5, 3, 1, 2, 2, 0, 1, 1, 1,
              2)

# Combined vectors into a data frame
our_class <- data.frame(first_name, last_name, email, distance)

View(our_class)
colnames(our_class)
```

***

**7. Change our_class so that *last_name* is in the column before *first_name*. Do this without creating a new data frame.**

```{r include=FALSE}
# Reorder columns so that last_name comes before first_name
our_class <- our_class[, c("last_name", "first_name", "email", "distance")]

# View order of column names
colnames(our_class)
```

***

**9. Use *our_class* to create a new data frame called *far_members* that contains the data for class members who travel at least 10 miles to class.**

```{r include=FALSE}
# Create the far_members data frame by subsetting our_class
far_members <- our_class[our_class$distance >= 10, ]

# View far_members data frame
View(far_members)
```

***

**10. Report the number of rows in the *far_members* data frame. Do not count by hand. Let R do the work.**

```{r include=FALSE}
# Report the number of rows in the far_members data frame
num_rows <- nrow(far_members)

num_rows
```

***

**11. The row names in *far_members* carried over from *our_class*. Change this so that they are in consecutive order. Write your code so it will work no matter how many rows are in *far_members*.**

```{r include=FALSE}
# Reset row names to be consecutive integers
rownames(far_members) <- 1:nrow(far_members)

# View far_members data frame
View(far_members)
```


***

**12. In one code statement, show all the information for the class member who travels the farthest to class. This should work even if we change the data in *our_class*.**

```{r include=FALSE}
# Show all information for the class member who travels the farthest

# our_class[which(our_class$distance)]
max_student <- our_class[which.max(our_class$distance), ]

max_student
```

***
