---
title: "EDRM 718 Exercise Set F"
Author: "Halle Prine"
output:
  html_document:
    df_print: paged
---

*Put your name as the author of this notebook in the YAML, then complete each of the items below. Make sure that your notebook is neat and organized and that all your code chunks run without error, even after clearing the environment. (Six points per item)*

```{r include=FALSE}
# The "here" library will let me read the data from an individual folder.
library(psych)
library(here)

# I read the data and put it in an object called "legislators".

students <- read.csv(here("Data", "hsb2.CSV"), header = TRUE, sep = ",") # Read the file
```

***

**These exercises refer to the *hsb2 data* that have been used in previous exercise sets.**

***

**1. Read in the *hsb2 data* and display the first 10 rows of the data set.**

```{r include=FALSE}
# The "here" library will let me read the data from an individual folder.
library(psych)
library(here)

# I read the data and put it in an object called "legislators".

students <- read.csv(here("Data", "hsb2.CSV"), header = TRUE, sep = ",") # Read the file

# Display the first 10 rows using Head and specify the 10 in the ()

head(students, 10)
```

***

**2. Extract the public school data and put these in a data frame called *public*. Extract the private school data and put these in a data frame called *private*.**

```{r include=FALSE}
# Extract public school data
public <- subset(students, schtyp == 1)

# Extract private school data
private <- subset(students, schtyp == 2)

# Display the first few rows of each dataset
head(public)
head(private)
```

***

**3. Put just test scores for the public schools in a matrix called *public_scores*, with each column representing a different subject area. Name the columns by the subject area. Do the same for private schools to create *private_scores*.**

```{r include=FALSE}
# Select only test score columns for public schools and convert to matrix
public_scores <- as.matrix(public[, c("read", "write", "math", "science", "socst")])
colnames(public_scores) <- c("Reading", "Writing", "Math", "Science", "Social Studies")

# Select only test score columns for private schools and convert to matrix
private_scores <- as.matrix(private[, c("read", "write", "math", "science", "socst")])
colnames(private_scores) <- c("Reading", "Writing", "Math", "Science", "Social Studies")


public_scores
private_scores
```

***

**4. Use a member of the apply family to create a vector *public_mean* and a vector *private_mean* that contain the means for the five tests for each type of school, respectively.**

```{r include=FALSE}
# Calculate mean test scores for public schools using apply
public_mean <- apply(public_scores, 2, mean)

# Calculate mean test scores for private schools using apply
private_mean <- apply(private_scores, 2, mean)

# Display the results
public_mean
private_mean
```

***

**5. The *colMeans()* function will calculate the means of columns of scores. Use this function to create a vector *public_mean_2* and a vector *private_mean_2* that contain the means for the five tests for each type of school, respectively.**

```{r include=FALSE}
# Calculate mean test scores for public schools using colMeans
public_mean_2 <- colMeans(public_scores)

# Calculate mean test scores for private schools using colMeans
private_mean_2 <- colMeans(private_scores)

# Display the results
public_mean_2
private_mean_2
```

***

**6. Use Boolean operators to determine if the class and type of *public_mean* is the same as that of *public_mean_2*. Also use a Boolean operator to determine if the class of *public_mean* is the same as the type of *public_mean*. Finally, use the *identical()* function to determine if *public_mean* and *public_mean_2* contain the same information.**

```{r include=FALSE}
# Check if the class and type of public_mean is the same as that of public_mean_2
class_public_mean <- (class(public_mean) == class(public_mean_2))  # Boolean comparison of class
class_public_mean # TRUE they are the same class

type_public_mean <- (typeof(public_mean) == typeof(public_mean_2))  # Boolean comparison of type
type_public_mean # TRUE they are the same type

# Check if the class of public_mean is the same as its type
class_type <- (class(public_mean) == typeof(public_mean))
class_type # FALSE it is NOT the same type 

# Check if public_mean and public_mean_2 contain the same information
identical_public_mean <- identical(public_mean, public_mean_2)
identical_public_mean # TRUE they both have the same information
```

***

**7. Create a vector *public_sd* and a vector *private_sd* that contain the standard deviations for the five tests for each type of school, respectively.**

```{r include=FALSE}
# Calculate standard deviations for public schools
public_sd <- apply(public_scores, 2, sd)

# Calculate standard deviations for private schools
private_sd <- apply(private_scores, 2, sd)

# Display the results
public_sd
private_sd
```

***

**8. Create a matrix of means and standard deviations called *public_stats* and another called *private_stats* for the public and private schools, respectively.**

```{r include=FALSE}
# Create matrix for public school statistics
public_stats <- rbind(public_mean, public_sd)

# Create matrix for private school statistics
private_stats <- rbind(private_mean, private_sd)

# Name the rows for clarity
rownames(public_stats) <- c("Mean", "SD")
rownames(private_stats) <- c("Mean", "SD")

# Display the matrices
public_stats
private_stats
```


***

**9. Create a well-labeled array named *hsb2_stats* that contains the means and standard deviations of public and private school students for the five tests in the hsb2 data. Each matrix should be named *Public* or *Private* and the statistics should be named as *Mean* and SD.**

```{r include=FALSE}

# Create the array with public and private statistics
hsb2_stats <- array(
  c(public_stats, private_stats),  # Combine public and private matrices
  dim = c(2, 5, 2),  # 2 rows (Mean, SD), 5 columns (Subjects), 2 matrices (Public, Private)
  dimnames = list(
    c("Mean", "SD"),  # Row names
    colnames(public_stats),  # Column names (Subjects)
    c("Public", "Private")  # Matrix names (School Type)
  )
)

# Display the array
hsb2_stats
```

***

**10. Create a list called *hsb2_info* that has two elements: a description of the array you created above and the actual array.**

```{r include=FALSE}
# Create a description of the array
description <- "This array contains the means and standard deviations of test scores for public and private school students across five subjects."

# Create the list with two elements
hsb2_info <- list(
  Description = description,
  Stats_Array = hsb2_stats
)

# Display the list
hsb2_info
```

***

**11. Illustrate how to display the description and the statistics contained in *hsb2_info* separately.**

```{r include=FALSE}
# Display the description
Description.1 <- hsb2_info$Description
Description.1
# OR
Description.2 <- hsb2_info[[1]]
Description.2

# Display the statistics array
stats_array.1 <- hsb2_info$Stats_Array
stats_array.1
# OR
stats_array.2 <- hsb2_info[[2]]
stats_array.2
```

***

**12. Extract the private school science mean from *hsb2_info*.**

```{r include=FALSE}
# Extract the private school science mean
private_science_mean <- hsb2_info$Stats_Array["Mean", "Science", "Private"]

# Display the result
private_science_mean
```

***
